Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-14 talk/icooolps2011/paper.tex
a slightly different approach to saving space: for the small examples, put code and trace next to each other. this makes it possible to use the larger font again.

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{relsize}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

@@

\usepackage[T1]{fontenc}
\usepackage[scaled=0.82]{beramono}

@@

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{lstlisting}
}

@@

The simplest example of constants are literal values, such as \texttt{1}.
However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code:

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
if x == 4:
    y = y + x
\end{lstlisting}
}

If the fragment is traced with $x_1$ being \texttt{4}, the following trace is
produced:
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$y_2$ = $y_1$ + $x_1$
\end{lstlisting}
}

In the trace above, the value of $x_1$ is statically known after the guard.
Remember that a guard is a runtime check. The above trace will run to
completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

@@

Let's make this more concrete. If we trace a call to the following function:
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}

We get a trace that looks like this:

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}

The hint indicates that \texttt{x} is likely a runtime constant and the JIT
should try to perform runtime specialization on it
in the code that follows.\footnote{For technical reasons the promote hint needs
to be written down slightly differently in the actual code.}  When just running
the code, the \texttt{promote} function has no
effect. When tracing, some extra work
is done. Let's assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning:

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_2$ = 9 + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 2)
$v_2$ = 5 + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

As an example, take the following class:

{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        self.y = self.compute() + val

    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}
}

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$x_1$ = $a_1$.x
$v_1$ = $x_1$ * 2
$v_2$ = $v_1$ + 1
$v_3$ = $v_2$ + $val_1$
$a_1$.y = $v_3$
\end{lstlisting}
}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        promote(self)
        self.y = self.compute() + val

    @elidable
    def compute(self):
        return self.x * 2 + 1
\end{lstlisting}
}

Now the trace will look like this:
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_1$ = compute($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{compute} function is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
{\smaller
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{wrapfig}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{relsize}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

@@

\usepackage[T1]{fontenc}
\usepackage[scaled=0.81]{beramono}

@@

\newcommand{\noop}{}

@@

{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{lstlisting}
}

@@

The simplest example of constants are literal values, such as \texttt{1}.
However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code on the left. If the fragment is traced with
$x_1$ being \texttt{4}, the trace on the left is produced:

\begin{minipage}[b]{0.5\linewidth}
\centering
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
if x == 4:
    y = y + x
\end{lstlisting}
}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$y_2$ = $y_1$ + $x_1$
\end{lstlisting}
}
\end{minipage}

In the trace, the value of $x_1$ is statically known after the guard.
Remember that a guard is a runtime check. The above trace will run to
completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

@@

Let's make this more concrete. If we trace a call to the function on the left, we get the trace on the right:

\begin{minipage}[b]{0.5\linewidth}
\centering
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}
\end{minipage}

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:

\begin{minipage}[b]{0.5\linewidth}
\centering
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}
\end{minipage}

The hint indicates that \texttt{x} is likely a runtime constant and the JIT
should try to perform runtime specialization on it
in the code that follows.\footnote{For technical reasons the promote hint needs
to be written down slightly differently in the actual code.}  When just running
the code, the \texttt{promote} function has no
effect. When tracing, some extra work
is done. Let's assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning.

@@

{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_2$ = 9 + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 2)
$v_2$ = 5 + $y_1$
return($v_2$)
\end{lstlisting}
}

@@

As an example, take the class on the left. Tracing the call \texttt{a.f(10)} of
some instance of \texttt{A} yields the trace on the right (note how the call to
\texttt{c} is inlined):

\begin{minipage}[t]{0.6\linewidth}
\centering
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        self.y = self.c() + val

    def c(self):
        return self.x * 2 + 1
\end{lstlisting}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[t]{0.4\linewidth}
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$x_1$ = $a_1$.x
$v_1$ = $x_1$ * 2
$v_2$ = $v_1$ + 1
$v_3$ = $v_2$ + $val_1$
$a_1$.y = $v_3$
\end{lstlisting}
\end{minipage}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{c}
is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
If the code in \texttt{c} should be constant-folded away, we would change the
class as follows:

\begin{minipage}[t]{0.6\linewidth}
\centering
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        promote(self)
        self.y = self.c() + val

    @elidable
    def c(self):
        return self.x * 2 + 1
\end{lstlisting}
\end{minipage}
\vline
\hspace{0.4cm}
\begin{minipage}[t]{0.4\linewidth}
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 
      0xb73984a8)
$v_1$ = c($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
\end{minipage}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{c} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{c} function is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}
